A Formal Verification Framework for Tezos Smart Contracts Based on Symbolic Execution

### 2024-01-18 papers/static-checker/ecoop2024/ecoop2024.tex
final cosmetics

\title{A Formal Verification Framework for Smart Contracts Based on Symbolic Execution} %TODO Please add

@@

Formal verification can guarantee the correctness of smart contracts.
To this end we have developed SCV, a formal
verification tool for smart contracts on the Tezos blockchain \cite{tezos-whitepaper}, written in
Michelson \cite{michelson}, although our design is widely applicable. SCV relies on
symbolic execution backed by an SMT solver to simulate the
implementation of the smart contract language, helping to detect 
subtle errors. In addition, SCV comes with a domain-specific language that
allows users to precisely specify contract properties.
% By interacting with an SMT solver, it can handle a wide range of properties.
It streamlines the process of reviewing requirements,
uncovering hidden errors, and validating user-defined properties.
SCV provides a comprehensive solution to
mitigate potential pitfalls in blockchain-based applications.

@@

Our contributions are as follows.
\begin{enumerate}
\item Design and implementation of the smart contract specification
  language SCV.
\item Design and implementation of a symbolic interpreter for a
  significant subset of Michelson.
\item Two case studies where we apply SCV to verify real life smart
  contracts deployed on the Tezos blockchain:
  \begin{itemize}
  \item USDtz, an implementation of the FA~1.2 standard for financial
    applications;
  \item the Kolibri oracle contract that injects currency exchange
    rates from an external system.
  \end{itemize}
\end{enumerate}

@@

To invoke a Michelson contract on the Tezos blockchain, the caller has
to provide tokens for gas and tokens to add to the contract's balance.
The gas tokens are used account for the cost of running the contract;
if contract execution runs out of gas, it is aborted before it can
perform any visible change. The use of the balance is up to the
discretion of the contract code.

@@

\section{A Domain Specific Language For Smart Contract Property Specification}
\label{sec:domain-specific-language}

% For Tezos users who may lack expertise in Michelson, there is a desire
% to verify smart contract code before interaction.
Our framework contains a domain-specific language (DSL) to specify
mathematical properties of smart contracts.
The DSL is designed to allow users to write formulas that align with
the capabilities of the underlying SMT solver (Z3).
The goal is to create a language
that is accessible and understandable for users who are
not Michelson experts, and guarantees formulas that the solver can handle effectively.

@@

\begin{figure}[tp]
\begin{align*}
\textbf{c} & ::= 
    \text{pre\_Balance}
   \Mid \text{post\_Balance}
   \Mid \text{Amount}
   \Mid \text{Sender}
   \Mid \text{Source} 
   \Mid \text{Now}
   \Mid \text{Level} \\
   &\Mid \text{Chain\_id}
   \Mid \text{Self\_address}\\
\textbf{t} &::= 
   \textbf{n}: \text{nat}
   \Mid\ \textbf{i}: \text{int}
   \Mid\  \textbf{s}: \text{string} 
   \Mid\  \textbf{b}: \text{bytes}
   \Mid\  \textbf{c}  
   \Mid\  \text{x} 
   \Mid\ \text{x}: \text{T} \\
   &\Mid\ \UNIT 
   \Mid\ \NEVER 
   \Mid\ \TRUE 
   \Mid\ \FALSE 
   \Mid\ \PAIR\ \text{t1 t2} \\
   &\Mid\ \LEFT\ \text{t}\  \text{T}
   \Mid\ \RIGHT\ \text{T}\ \text{t}
   \Mid\ \SOME\ \text{t}
   \Mid\ \NONE\ \text{T} 
   \Mid\ \text{\{t ; ... \}}: \text{list T}
   \Mid\ \text{\{t ; ... \}}: \text{set T} \\
   &\Mid\ \text{\{ Elt t1 t2 ; ... \}}: \text{map T U}
   \Mid\ \text{\{ Elt t1 t2 ; ... \}}: \text{big\_map T U} \\
\textbf{e} & :: = \textbf{t}  \Mid \textbf{e} \text{ + } \textbf{e}\Mid \textbf{e} \text{ - } \textbf{e}  \Mid \textbf{e} \text{ * } \textbf{e} \Mid \textbf{e} \text{ / } \textbf{e} \Mid \textbf{e} \text{ \% } \textbf{e} \Mid \textbf{e} \text{ rem } \textbf{e} \\
& 
 \Mid \text{get\_map} (\textbf{e1}, \textbf{e2}) 
 \Mid \text{get\_big\_map} (\textbf{e1}, \textbf{e2})  \\
 &\Mid \text{get\_contract} (\textbf{e1}, \textbf{e2})   
 \Mid \text{get\_substring} (\textbf{e1}, \textbf{e2}, \textbf{e3})   \\
& \Mid \text{abs} (\textbf{e})  
 \Mid \text{to\_int} (\textbf{e})  
 \Mid \text{len} (\textbf{e})
 \Mid \text{cons} (\textbf{e}, \textbf{e})
 \Mid \text{size} (\textbf{e})
\Mid \textbf{e} \text{ $\wedge$ } \textbf{e}  \\
\textbf{p} & :: = \textbf{e} \text{ = } \textbf{e}
\Mid \textbf{e} \text{ != } \textbf{e}
\Mid \textbf{e} \text{ > } \textbf{e}
\Mid \textbf{e} \text{ >= } \textbf{e}
\Mid \textbf{e} \text{ < } \textbf{e}
\Mid \textbf{e} \text{ <= } \textbf{e} \\
& \Mid \text{transfer\_token} (\textbf{e1}, \textbf{e2}, \textbf{e3})  
 \Mid \text{is\_nat} (\textbf{e})  
 \Mid \text{contain} (\textbf{e1}, \textbf{e2}) \\
& \Mid \textbf{p} \text{ and } \textbf{p}\Mid \textbf{p} \text{ or } \textbf{p}  \Mid \textbf{p} \text{ nor } \textbf{p} \Mid  \text{ not } \textbf{p} \Mid \textbf{p} \text{ => } \textbf{p} \\
\end{align*}
\caption{Syntax of terms, t and predicates, p}
\label{fig:syntax-terms-predicates}
\end{figure}

\begin{lstlisting}[float=tp,captionpos=b,caption={Smart contract property specification for a simple contract},label={lst:smart-contract-property-specification},numbers=left]
mcontract name = spec
  code := {...}
  input := t
  output := t'
  pre-condition := cd
  post-condition := cd'
\end{lstlisting}

@@

Formally, the pre- and postcondition instances are formulas in
first-order logic that can be handled by an SMT solver (in our case,
Z3). The specification language is built from
predicates. Figure~\ref{fig:syntax-terms-predicates} defines the
syntax of predicates. Formulas that are used in assertions or added to solvers are terms of sort Boolean. Otherwise, terms of
Boolean and non-Boolean sort may be mixed in any combination where
sorts match up. Several functions are predefined, such as the function
\text{get\_map (e1, e2)} for accessing an element in a map. This
function returns \text{None} if there is no key \text{e1} in the map
\text{e1}, or the corresponding value of the key otherwise. Another
function, \text{get\_contract (e1, e2)}, casts the address \text{e2}
to a real contract on the Tezos blockchain. It returns \text{None} if
there is no existing contract on the Tezos block chain with an
entrypoint of type \text{e1} associated with the address \text{e2}, or
it returns the contract otherwise.

@@

The specification continues by defining
pre- and postconditions, where the precondition states the assumption
that the integer \lstinline/x/ is greater than or equal to  \lstinline/y/. The postcondition asserts that the result
\lstinline/z/ is greater than or equal to the substract of
\lstinline/x/ and \lstinline/y/, which is greater or equal to
zero. The symbol \lstinline/&&/ stands for logical predicate
conjunction.

@@

\begin{lstlisting}[float,captionpos=b,caption={Multiple entrypoint specification syntax (option 1)},label={lst:multiple-entrypoint-specification-1},numbers=left][h]
mcontract name = spec
   entrypoint %a 
      code := {c1}
      input := t1
      output := t1'
      pre-condition := cd1
      post-condition := cd1'
   entrypoint %b 
      ...
   <entrypoint relations>
\end{lstlisting}

\begin{lstlisting}[float,captionpos=b,caption={Multiple entrypoint specification syntax (option 2)},label={lst:multiple-entrypoint-specification-2},numbers=left][h]
mcontract name = spec
   storage := t1
   entrypoint %a 
      code := {c1}
      parameter := t2
      pre-condition := cd1
      post-condition := cd1'
   entrypoint %b 
      ...
   <entrypoint relations>
\end{lstlisting}

@@

\subsection{Entrypoint relations}
\label{sec:entrypoint-relations}
For a smart contract on the blockchain, the order of smart contract
calls can be crucial. We specify ordering constraints among calls with
an entrypoint relation. An entrypoint relation can be specified as
\lstinline/%a -> %b with cd/, which starts with the name
\lstinline/%a/ of the entrypoint and is followed by the keyword
\lstinline/with/ in case conditions are specified as predicates. The
following symbol \lstinline/->/ indicates that from the entrypoint
that satisfies the condition \lstinline/cd/, it is possible  to call
the entrypoint \lstinline/%b/. One smart contract specification can
contain several entrypoint relations concatenated by
\lstinline/|/. There is a special entrypoint named \lstinline/create/
that indicates the contract state after deployment.

@@

% \paragraph{Dealing with loop formalization}
% \label{sec:dealing-with-loop}
Our execution model formalizes loops symbolically using the \FOLD\ and
\FMAP\ functions (see
Section~\ref{sec:symbolic-execution-model}). This approach has the
advantage of allowing us to convert these loops into formulas
compatible with certain SMT solvers, such as Z3, which support these
functions. Specifically, Z3 provides \lstinline/seq.map fn s/, a map
function on sequence \lstinline/s/, and
\lstinline/(seq.foldl fn b s)/, a fold function with an initial value \lstinline/b/ on sequence
\lstinline/s/. These functions are supported for checking assertions
that contain them. Therefore, we can directly convert the symbolic
results of loops into formulas in Z3. 
\subsection{Static checker}
\label{sec:static-checker}
The static checker now performs the verification process by employing symbolic execution against the property specification in the DSL. It utilizes the results from the previous phases, including symbolic execution, Z3 formulas representing pre and post conditions, and path conditions.

@@

\subsubsection{Life-cycle check}
\label{sec:life-cycle-check}
An entrypoint relation can be defined as \lstinline/(%a -> %b) with cd/, indicating that after invoking the entrypoint \lstinline/%a/, it is permissible to call the entrypoint \lstinline/%b/ if the condition \lstinline/cd/ is satisfied. The presence of the keyword \lstinline/not/ at the beginning signals an impossibility. Note that \lstinline/with cd/ may be omitted when no condition is needed.

Let $S_a$ represent the set of all reachable states resulting from the execution of the call to the entrypoint \lstinline/%a/. Each $s_{i}$ in $S_a$ is associated with a path condition $\Theta_i$ and is represented as a stack containing a single element—a pair consisting of an operation list ($\VOPERATIONLIST_i$) and the storage $\VSTORAGE_i$. For each $s_{i}$, symbolic execution is performed on the entrypoint \lstinline/%b/ using input as a pair of the parameter and the storage $\VSTORAGE_i$, constrained by $\Theta_i$ and the condition \lstinline/cd/. In the execution state space, a state $s_{j}$ with the path condition $\Theta_j$ is considered reachable if ($\Theta_i$ $\land$ \lstinline/cd/ $\land$ $\Theta_j$) is satisfiable, and $\neg$ ($\Theta_i$ $\land$ \lstinline/cd/ $\rightarrow$ $\Theta_j$) is unsatisfiable. Let $S_b$ denote the set of these reachable states. If $S_b$ is empty or contains only failwith states, then it is impossible to invoke the entrypoint \lstinline/%b/ from the state $s_{i}$. Otherwise, it is considered a success. Let $R$ be the set containing all such states $s_{i}$ in $S_a$ where it is successful to call the entrypoint \lstinline/%b/ with the condition \lstinline/cd/. If $R$ is non-empty, then \lstinline/(%a -> %b) with cd/ holds. Otherwise, it is false, signifying \lstinline/not (%a  -> %b) with cd/.

@@

We already learned about Michelson, the low-level smart contract
language for the Tezos blockchain, and its stack-based nature in
Section~\ref{sec:background}.  Here we consider the changes needed for
symbolic execution of Michelson programs.

>>>>

\title{A Formal Verification Framework for Tezos Smart Contracts Based on Symbolic Execution} %TODO Please add

@@

Formal verification can guarantee the correctness of smart contracts.
To this end we have developed SCV, a formal
verification tool for smart contracts on the Tezos blockchain \cite{tezos-whitepaper}, written in
Michelson \cite{michelson}. SCV relies on
symbolic execution backed by an SMT solver to simulate the
implementation of the smart contract language, helping to detect 
subtle errors. In addition, SCV comes with a domain-specific language that
allows users to precisely specify contract properties.
% By interacting with an SMT solver, it can handle a wide range of properties.
It streamlines the process of reviewing requirements,
uncovering hidden errors, and validating user-defined properties.
SCV provides a comprehensive solution to
mitigate potential pitfalls in blockchain-based applications.

@@

Our contributions are as follows and supported by the upcoming
artifact submission.
\begin{enumerate}
\item Design and implementation of the smart contract specification
  language SCV.
\item Design and implementation of a symbolic interpreter for a
  significant subset of Michelson.
\item Two case studies that verify smart
  contracts deployed on the Tezos blockchain with SCV:
  \begin{itemize}
  \item USDtz, an implementation of the FA~1.2 standard for financial
    applications;
  \item Kolibri, an oracle contract that injects currency exchange
    rates from an external system.
  \end{itemize}
\end{enumerate}

@@

To invoke a Michelson contract on the Tezos blockchain, the caller has
to provide tokens for gas and tokens to add to the contract's balance.
The gas tokens are used account for the cost of running the contract;
if contract execution runs out of gas, it is aborted before it can
perform any visible change. Use of the balance is up to the contract code.

@@

\section{SCV: A Domain Specific Language For Smart Contract Property Specification}
\label{sec:domain-specific-language}

% For Tezos users who may lack expertise in Michelson, there is a desire
% to verify smart contract code before interaction.
Our framework SCV contains a domain-specific language to specify
mathematical properties of smart contracts.
SCV is designed to allow users to write formulas that align with
the capabilities of the underlying SMT solver (Z3).
The goal is to create a language
that is accessible and understandable for users who are
not Michelson experts, and guarantees formulas that the solver can handle effectively.

@@

\begin{figure}[tp]
\begin{align*}
\textbf{c} & ::= 
    \text{pre\_Balance}
   \Mid \text{post\_Balance}
   \Mid \text{Amount}
   \Mid \text{Sender}
   \Mid \text{Source} 
   \Mid \text{Now}
   \Mid \text{Level} \\
   &\Mid \text{Chain\_id}
   \Mid \text{Self\_address}\\
\textbf{t} &::= 
   \textbf{n}: \text{nat}
   \Mid\ \textbf{i}: \text{int}
   \Mid\  \textbf{s}: \text{string} 
   \Mid\  \textbf{b}: \text{bytes}
   \Mid\  \textbf{c}  
   \Mid\  \text{x} 
   \Mid\ \text{x}: \text{T} \\
   &\Mid\ \UNIT 
   \Mid\ \NEVER 
   \Mid\ \TRUE 
   \Mid\ \FALSE 
   \Mid\ \PAIR\ \text{t1 t2} \\
   &\Mid\ \LEFT\ \text{t}\  \text{T}
   \Mid\ \RIGHT\ \text{T}\ \text{t}
   \Mid\ \SOME\ \text{t}
   \Mid\ \NONE\ \text{T} 
   \Mid\ \text{\{t ; ... \}}: \text{list T}
   \Mid\ \text{\{t ; ... \}}: \text{set T} \\
   &\Mid\ \text{\{ Elt t1 t2 ; ... \}}: \text{map T U}
   \Mid\ \text{\{ Elt t1 t2 ; ... \}}: \text{big\_map T U} \\
\textbf{e} & :: = \textbf{t}  \Mid \textbf{e} \text{ + } \textbf{e}\Mid \textbf{e} \text{ - } \textbf{e}  \Mid \textbf{e} \text{ * } \textbf{e} \Mid \textbf{e} \text{ / } \textbf{e} \Mid \textbf{e} \text{ \% } \textbf{e} \Mid \textbf{e} \text{ rem } \textbf{e} \\
& 
 \Mid \text{get\_map} (\textbf{e1}, \textbf{e2}) 
 \Mid \text{get\_big\_map} (\textbf{e1}, \textbf{e2})  \\
 &\Mid \text{get\_contract} (T, \textbf{e})   
 \Mid \text{get\_substring} (\textbf{e1}, \textbf{e2}, \textbf{e3})   \\
& \Mid \text{abs} (\textbf{e})  
 \Mid \text{to\_int} (\textbf{e})  
 \Mid \text{len} (\textbf{e})
 \Mid \text{cons} (\textbf{e}, \textbf{e})
 \Mid \text{size} (\textbf{e})
\Mid \textbf{e} \text{ $\wedge$ } \textbf{e}  \\
\textbf{p} & :: = \textbf{e} \text{ = } \textbf{e}
\Mid \textbf{e} \text{ != } \textbf{e}
\Mid \textbf{e} \text{ > } \textbf{e}
\Mid \textbf{e} \text{ >= } \textbf{e}
\Mid \textbf{e} \text{ < } \textbf{e}
\Mid \textbf{e} \text{ <= } \textbf{e} \\
& \Mid \text{transfer\_token} (\textbf{e1}, \textbf{e2}, \textbf{e3})  
 \Mid \text{is\_nat} (\textbf{e})  
 \Mid \text{contain} (\textbf{e1}, \textbf{e2}) \\
& \Mid \textbf{p} \text{ \&\& } \textbf{p}\Mid \textbf{p} \text{ or } \textbf{p}  
% \Mid \textbf{p} \text{nor } \textbf{p} 
\Mid  \text{ not } \textbf{p} \Mid \textbf{p} \text{ => } \textbf{p} \\
\end{align*}
\caption{Syntax of constants, \textbf{c}, terms, \textbf{t}, expressions, \textbf{e}, and predicates, \textbf{p}}
\label{fig:syntax-terms-predicates}
\end{figure}

\begin{lstlisting}[float=tp,captionpos=b,caption={Smart contract property specification for a simple contract},label={lst:smart-contract-property-specification},numbers=left]
mcontract name = spec
  code := {...}
  input := t
  output := t'
  pre-condition := p
  post-condition := p'
\end{lstlisting}

@@

Formally, the pre- and postcondition instances are formulas in
first-order logic that can be handled by an SMT solver (in our case,
Z3). The specification language is built from
predicates. Figure~\ref{fig:syntax-terms-predicates} defines the
syntax of predicates. Formulas that are used in assertions or added to solvers are terms of sort Boolean. Otherwise, terms of
Boolean and non-Boolean sort may be mixed in any combination where
sorts match up. Several functions are predefined, such as the function
\text{get\_map (e1, e2)} for accessing an element in a map. This
function returns \text{None} if there is no key \text{e1} in the map
\text{e1}, or \text{Some(v)} if $v$ is the value mapped by the key. Another
function, \text{get\_contract (T, e)}, casts the address \text{e}
to a contract on the Tezos blockchain. It returns \text{None} if
there is no contract on the Tezos blockchain with an
entrypoint of type \text{T} associated with the address \text{e}, or
it returns the contract in the form \text{Some(c)}.

@@

The specification continues by defining
pre- and postconditions, where the precondition states the assumption
that the integer \lstinline/x/ is greater than or equal to  \lstinline/y/. The postcondition asserts that the result
\lstinline/z/ is greater than or equal to the substract of
\lstinline/x/ and \lstinline/y/, which is greater or equal to
zero. The symbol \lstinline/&&/ stands for conjunction.

@@

\begin{lstlisting}[float,captionpos=b,caption={Multiple entrypoint specification syntax (option 1)},label={lst:multiple-entrypoint-specification-1},numbers=left][h]
mcontract name = spec
   entrypoint %a 
      code := {c1}
      input := t1
      output := t1'
      pre-condition := p1
      post-condition := p1'
   entrypoint %b 
      ...
   <entrypoint relations>
\end{lstlisting}

\begin{lstlisting}[float,captionpos=b,caption={Multiple entrypoint specification syntax (option 2)},label={lst:multiple-entrypoint-specification-2},numbers=left][h]
mcontract name = spec
   storage := t1
   entrypoint %a 
      code := {c1}
      parameter := t2
      pre-condition := p1
      post-condition := p1'
   entrypoint %b 
      ...
   <entrypoint relations>
\end{lstlisting}

@@

\subsection{Entrypoint relations}
\label{sec:entrypoint-relations}
For a smart contract on the blockchain, the order of smart contract
calls can be crucial. We specify ordering constraints among calls with
an entrypoint relation. An entrypoint relation can be specified as
\lstinline/%a -> %b with p/, which starts with the name
\lstinline/%a/ of the entrypoint and is followed by the keyword
\lstinline/with/ in case conditions are specified as predicates. The
following symbol \lstinline/->/ indicates that from the entrypoint
that satisfies the condition \lstinline/p/, it is possible  to call
the entrypoint \lstinline/%b/. One smart contract specification can
contain several entrypoint relations concatenated by
\lstinline/|/. There is a special entrypoint named \lstinline/create/
that indicates the contract state after deployment.

@@

% \paragraph{Dealing with loop formalization}
% \label{sec:dealing-with-loop}
Our execution model formalizes loops symbolically using the \FOLD\ and
\FMAP\ functions (see
Section~\ref{sec:symbolic-execution-model}). This approach has the
advantage of allowing us to convert these loops into formulas
compatible with certain SMT solvers, such as Z3, which support these
functions. Specifically, Z3 provides \lstinline/seq.map fn s/, a map
function on sequence \lstinline/s/, and
\lstinline/(seq.foldl fn b s)/, a fold function with an initial value \lstinline/b/ on sequence
\lstinline/s/. These functions are supported for checking assertions
that contain them. Therefore, we can directly convert the symbolic
results of loops into formulas in Z3. 
\subsection{Static checker}
\label{sec:static-checker}
The static checker now performs the verification process by employing symbolic execution against the property specification in SCV. It utilizes the results from the previous phases, including symbolic execution, Z3 formulas representing pre and post conditions, and path conditions.

@@

\subsubsection{Life-cycle check}
\label{sec:life-cycle-check}
An entrypoint relation can be defined as \lstinline/(%a -> %b) with p/, indicating that after invoking the entrypoint \lstinline/%a/, it is permissible to call the entrypoint \lstinline/%b/ if the condition \lstinline/p/ is satisfied. The presence of the keyword \lstinline/not/ at the beginning signals an impossibility. Note that \lstinline/with p/ may be omitted when no condition is needed.

Let $S_a$ represent the set of all reachable states resulting from the execution of the call to the entrypoint \lstinline/%a/. Each $s_{i}$ in $S_a$ is associated with a path condition $\Theta_i$ and is represented as a stack containing a single element—a pair consisting of an operation list ($\VOPERATIONLIST_i$) and the storage $\VSTORAGE_i$. For each $s_{i}$, symbolic execution is performed on the entrypoint \lstinline/%b/ using input as a pair of the parameter and the storage $\VSTORAGE_i$, constrained by $\Theta_i$ and the condition \lstinline/p/. In the execution state space, a state $s_{j}$ with the path condition $\Theta_j$ is considered reachable if ($\Theta_i$ $\land$ \lstinline/p/ $\land$ $\Theta_j$) is satisfiable, and $\neg$ ($\Theta_i$ $\land$ \lstinline/p/ $\rightarrow$ $\Theta_j$) is unsatisfiable. Let $S_b$ denote the set of these reachable states. If $S_b$ is empty or contains only failwith states, then it is impossible to invoke the entrypoint \lstinline/%b/ from the state $s_{i}$. Otherwise, it is considered a success. Let $R$ be the set containing all such states $s_{i}$ in $S_a$ where it is successful to call the entrypoint \lstinline/%b/ with the condition \lstinline/p/. If $R$ is non-empty, then \lstinline/(%a -> %b) with p/ holds. Otherwise, it is false, signifying \lstinline/not (%a  -> %b) with p/.

@@

We already learned about Michelson, the low-level smart contract
language for the Tezos blockchain, and its stack-based nature in
Section~\ref{sec:background}.  Here we consider the changes needed for
symbolic execution of Michelson programs. We only discuss symbolic
execution for some select instructions. The
appendix~\ref{sec:concl-sect-append} specifies further instructions,
but may be skipped by the reviewers.
